cubble: An R Package for Structuring Spatio-temporal Data

### 2022-04-27 1-intro.Rmd
Introduction sections changes and comments

Spatio-temporal data has a spatial component referring to the location of each observation, a temporal component that is recorded at regular or irregular time intervals, and may have multiple variables measured at each spatial and temporal value. With spatio-temporal data, one can fix the time to explore the spatial distribution, fix the spatial location to explore temporal trends, or dynamically explore the space and time simultaneously. Underlying all these tasks is a data representation that enables the spatio-temporal data to be sliced and diced differently.

The SpatioTemproal CRAN task view [@ctvspatiotemporal] has a section *Representing data* that lists existing spatio-temporal data representations in \proglang{R}. Among them, @spacetime proposes four spatio-temporal layouts: full grid layout, sparse grid layout, irregular layout, and trajectory. These layouts are implemented in the \pkg{spacetime} [@spacetime] package with the underlying spatial and temporal classes \pkg{sp} [@sp] and \pkg{xts} [@xts]. The \pkg{stars} [@stars] package has been a recent implementation to handle both raster and vector data using spatio-temporal arrays. It interfaces with \pkg{sf} [@sf], which adopts the simple features representation of spatial data, and \pkg{tidyverse} [@tidyverse], a suite of tools to wrangle and visualise data in \proglang{R}.

Still, the data representation for spatio-temporal data can be further improved and there are two reasons for this. Firstly, the raw data sourced in the wild is less often presented in any one of the representations above, and fitting the raw data into a data object can sometimes be difficult. More often, spatio-temporal data are collected in separate 2D tables and analysts need to assemble them into a whole piece before exploring the data. Examples of components of spatio-temporal data can be 1) map data recording the shape of a collection of areas of interest; 2) geo-scientific data recording the longitude and latitude coordinates of locations in the area along with other metadata related to the location, and; 3) temporal variables indexed by both location and time.

@@

The other reason is how tidy data concepts [@tidydata] should be applied to spatio-temporal data. Tidy data should be structured into 1) one row per observation, 2) one column per variable, and 3) one type of data per table. The long form data is preferred over wide data given the downstream software i.e. \pkg{dplyr} [@dplyr] and \pkg{ggplot2} [@ggplot2] for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygons for daily or sub-daily data over years, which are extensively collected and handled in time series analysis. This poses the question of how to arrange spatial and temporal variables in a way that would make analysing spatio-temporal data easier.

This paper presents a new \proglang{R} package, \pkg{cubble} which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while being kept synchronised. Among the four spacetime layouts in @spacetime, \pkg{cubble} can be applied to full grid layout, sparse grid layout, or irregular layout, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link].

The rest of the paper is divided as follows: Section \@ref(cube) introduces the proposed cube structure as a way to conceptualise multivariate spatio-temporal data. Section \@ref(cubble) presents the main design and functionality of \pkg{cubble}. Section \@ref(others) explains how cubble deals with more advanced considerations, including data with the hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section \@ref(examples) uses Australian weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example of how \pkg{cubble} handles NetCDF data is also provided. Section \@ref(conclude) concludes the paper.

>>>>

Spatio-temporal data has a spatial component referring to the location of each observation and a temporal component that is recorded at regular or irregular time intervals. It may also include multiple variables measured at each spatial and temporal values. With spatio-temporal data, one can fix the time to explore the spatial features of the data, fix the spatial location/s to explore temporal aspects, or dynamically explore the space and time simultaneously. 
<!-- Underlying all these tasks, there is a data representation that enables the spatio-temporal data to be sliced and diced differently.  -->
In order to computationally explore the spatial, temporal and spatio-temporal faces of such data, the data needs to be stored and represented under a specific data object that allows the user to query, group and dissect all the data faces.**

The SpatioTemporal CRAN task view [@ctvspatiotemporal] gathers information about R packages designed for spatio-temporal data and it has a section on *Representing data* that lists existing spatio-temporal data representations used in \proglang{R}. Among them, @spacetime proposes four spatio-temporal layouts: full grid layout, sparse grid layout, irregular layout, and trajectory. These layouts are implemented in the \pkg{spacetime} [@spacetime] package via the underlying spatial and temporal classes \pkg{sp} [@sp] and \pkg{xts} [@xts]. The \pkg{stars} [@stars] package has been a recent implementation to handle both raster and vector data using spatio-temporal arrays. It interfaces with \pkg{sf} [@sf], which adopts the simple features representation of spatial data, and \pkg{tidyverse} [@tidyverse], a suite of tools to wrangle and visualise data in \proglang{R}.

**The paragraph above needs a bit more flesh**

Still, the data representation for spatio-temporal data can be further improved and there are two reasons for this. Firstly, the raw data sourced in the wild is less often presented in any one of the representations above (Which representations - in the previous paragraph there are more than two?), and fitting the raw data into a data object can sometimes be difficult. More often, spatio-temporal data are collected in separate 2D tables and analysts need to assemble them into a whole piece before exploring the data. Examples of components of spatio-temporal data can be 1) map data recording the shape of a collection of areas of interest; 2) geo-scientific data recording the longitude and latitude coordinates of locations in the area along with other metadata related to the location, and; 3) temporal variables indexed by both location and time.

**I would aligned the examples of spatio-temporal data with those discuss in the st books. See Paula Moraga for example**

@@

The other reason is about tidy data concepts [@tidydata] and how they should be applied to spatio-temporal data. Tidy data should be structured into 1) one row per observation, 2) one column per variable, and 3) one type of data per table. The long form data is preferred over wide data given the downstream software i.e. \pkg{dplyr} [@dplyr] and \pkg{ggplot2} [@ggplot2] for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygons for hourly, daily or sub-daily periods over years, which are extensively collected and handled in time series analysis. This poses the question of how to arrange spatial and temporal variables in a way that would make data wrangling, visualizing and analysing spatio-temporal data easier.

This paper presents a new \proglang{R} package, \pkg{cubble} which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately or combined while being kept synchronised. Among the four spacetime layouts in @spacetime, \pkg{cubble} can be applied to full di layout, sparse grid layout, or irregular layout, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link]. **At this point what is cubble is adding with respect the other packages it is not  clear to me**

The rest of the paper is organized as follows: Section \@ref(cube) introduces the proposed cube structure as a way to conceptualise multivariate spatio-temporal data. Section \@ref(cubble) presents the main design and functionality of \pkg{cubble}. Section \@ref(others) explains how cubble deals with more advanced considerations, including data with hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section \@ref(examples) uses Australian weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example of how \pkg{cubble} handles NetCDF data is also provided. Section \@ref(conclude) concludes the paper.

### 2022-04-27 arxiv/cubble-arxiv.tex
Introduction sections changes and comments

Spatio-temporal data has a spatial component referring to the location of each observation, a temporal component that is recorded at regular or irregular time intervals, and may have multiple variables measured at each spatial and temporal value. With spatio-temporal data, one can fix the time to explore the spatial distribution, fix the spatial location to explore temporal trends, or dynamically explore the space and time simultaneously. Underlying all these tasks is a data representation that enables the spatio-temporal data to be sliced and diced differently.

The SpatioTemproal CRAN task view (\protect\hyperlink{ref-ctvspatiotemporal}{Edzer Pebesma 2022}) has a section \emph{Representing data} that lists existing spatio-temporal data representations in \proglang{R}. Among them, \protect\hyperlink{ref-spacetime}{E. Pebesma} (\protect\hyperlink{ref-spacetime}{2012}) proposes four spatio-temporal layouts: full grid layout, sparse grid layout, irregular layout, and trajectory. These layouts are implemented in the \pkg{spacetime} (\protect\hyperlink{ref-spacetime}{E. Pebesma 2012}) package with the underlying spatial and temporal classes \pkg{sp} (\protect\hyperlink{ref-sp}{E. Pebesma and Bivand 2005}) and \pkg{xts} (\protect\hyperlink{ref-xts}{Ryan and Ulrich 2020}). The \pkg{stars} (\protect\hyperlink{ref-stars}{E. Pebesma 2021}) package has been a recent implementation to handle both raster and vector data using spatio-temporal arrays. It interfaces with \pkg{sf} (\protect\hyperlink{ref-sf}{E. J. Pebesma 2018}), which adopts the simple features representation of spatial data, and \pkg{tidyverse} (\protect\hyperlink{ref-tidyverse}{Wickham et al. 2019}), a suite of tools to wrangle and visualise data in \proglang{R}.

Still, the data representation for spatio-temporal data can be further improved and there are two reasons for this. Firstly, the raw data sourced in the wild is less often presented in any one of the representations above, and fitting the raw data into a data object can sometimes be difficult. More often, spatio-temporal data are collected in separate 2D tables and analysts need to assemble them into a whole piece before exploring the data. Examples of components of spatio-temporal data can be 1) map data recording the shape of a collection of areas of interest; 2) geo-scientific data recording the longitude and latitude coordinates of locations in the area along with other metadata related to the location, and; 3) temporal variables indexed by both location and time.

The other reason is how tidy data concepts (\protect\hyperlink{ref-tidydata}{Wickham 2014}) should be applied to spatio-temporal data. Tidy data should be structured into 1) one row per observation, 2) one column per variable, and 3) one type of data per table. The long form data is preferred over wide data given the downstream software i.e.~\pkg{dplyr} (\protect\hyperlink{ref-dplyr}{Wickham et al. 2022}) and \pkg{ggplot2} (\protect\hyperlink{ref-ggplot2}{Wickham 2016}) for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygons for daily or sub-daily data over years, which are extensively collected and handled in time series analysis. This poses the question of how to arrange spatial and temporal variables in a way that would make analysing spatio-temporal data easier.

This paper presents a new \proglang{R} package, \pkg{cubble} which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while being kept synchronised. Among the four spacetime layouts in \protect\hyperlink{ref-spacetime}{E. Pebesma} (\protect\hyperlink{ref-spacetime}{2012}), \pkg{cubble} can be applied to full grid layout, sparse grid layout, or irregular layout, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}.

The rest of the paper is divided as follows: Section \ref{cube} introduces the proposed cube structure as a way to conceptualise multivariate spatio-temporal data. Section \ref{cubble} presents the main design and functionality of \pkg{cubble}. Section \ref{others} explains how cubble deals with more advanced considerations, including data with the hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section \ref{examples} uses Australian weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example of how \pkg{cubble} handles NetCDF data is also provided. Section \ref{conclude} concludes the paper.

>>>>

Spatio-temporal data has a spatial component referring to the location of each observation and a temporal component that is recorded at regular or irregular time intervals. It may also include multiple variables measured at each spatial and temporal values. With spatio-temporal data, one can fix the time to explore the spatial features of the data, fix the spatial location/s to explore temporal aspects, or dynamically explore the space and time simultaneously.
In order to computationally explore the spatial, temporal and spatio-temporal faces of such data, the data needs to be stored and represented under a specific data object that allows the user to query, group and dissect all the data faces.**

The SpatioTemporal CRAN task view (\protect\hyperlink{ref-ctvspatiotemporal}{Edzer Pebesma 2022}) gathers information about R packages designed for spatio-temporal data and it has a section on \emph{Representing data} that lists existing spatio-temporal data representations used in \proglang{R}. Among them, \protect\hyperlink{ref-spacetime}{E. Pebesma} (\protect\hyperlink{ref-spacetime}{2012}) proposes four spatio-temporal layouts: full grid layout, sparse grid layout, irregular layout, and trajectory. These layouts are implemented in the \pkg{spacetime} (\protect\hyperlink{ref-spacetime}{E. Pebesma 2012}) package via the underlying spatial and temporal classes \pkg{sp} (\protect\hyperlink{ref-sp}{E. Pebesma and Bivand 2005}) and \pkg{xts} (\protect\hyperlink{ref-xts}{Ryan and Ulrich 2020}). The \pkg{stars} (\protect\hyperlink{ref-stars}{E. Pebesma 2021}) package has been a recent implementation to handle both raster and vector data using spatio-temporal arrays. It interfaces with \pkg{sf} (\protect\hyperlink{ref-sf}{E. J. Pebesma 2018}), which adopts the simple features representation of spatial data, and \pkg{tidyverse} (\protect\hyperlink{ref-tidyverse}{Wickham et al. 2019}), a suite of tools to wrangle and visualise data in \proglang{R}.

\textbf{The paragraph above needs a bit more flesh}

Still, the data representation for spatio-temporal data can be further improved and there are two reasons for this. Firstly, the raw data sourced in the wild is less often presented in any one of the representations above (Which representations - in the previous paragraph there are more than two?), and fitting the raw data into a data object can sometimes be difficult. More often, spatio-temporal data are collected in separate 2D tables and analysts need to assemble them into a whole piece before exploring the data. Examples of components of spatio-temporal data can be 1) map data recording the shape of a collection of areas of interest; 2) geo-scientific data recording the longitude and latitude coordinates of locations in the area along with other metadata related to the location, and; 3) temporal variables indexed by both location and time.

\textbf{I would aligned the examples of spatio-temporal data with those discuss in the st books. See Paula Moraga for example}

The other reason is about tidy data concepts (\protect\hyperlink{ref-tidydata}{Wickham 2014}) and how they should be applied to spatio-temporal data. Tidy data should be structured into 1) one row per observation, 2) one column per variable, and 3) one type of data per table. The long form data is preferred over wide data given the downstream software i.e.~\pkg{dplyr} (\protect\hyperlink{ref-dplyr}{Wickham et al. 2022}) and \pkg{ggplot2} (\protect\hyperlink{ref-ggplot2}{Wickham 2016}) for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygons for hourly, daily or sub-daily periods over years, which are extensively collected and handled in time series analysis. This poses the question of how to arrange spatial and temporal variables in a way that would make data wrangling, visualizing and analysing spatio-temporal data easier.

This paper presents a new \proglang{R} package, \pkg{cubble} which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately or combined while being kept synchronised. Among the four spacetime layouts in \protect\hyperlink{ref-spacetime}{E. Pebesma} (\protect\hyperlink{ref-spacetime}{2012}), \pkg{cubble} can be applied to full di layout, sparse grid layout, or irregular layout, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}. \textbf{At this point what is cubble is adding with respect the other packages it is not clear to me}

The rest of the paper is organized as follows: Section \ref{cube} introduces the proposed cube structure as a way to conceptualise multivariate spatio-temporal data. Section \ref{cubble} presents the main design and functionality of \pkg{cubble}. Section \ref{others} explains how cubble deals with more advanced considerations, including data with hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section \ref{examples} uses Australian weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example of how \pkg{cubble} handles NetCDF data is also provided. Section \ref{conclude} concludes the paper.
